Nested Sampling with Multiple Objective Functions

### 2018-06-16 doc/ms.tex
A bit of writing on a Saturday afternoon

The combination of the reps makes use of the (approximate?) unbiasedness
property of Nested Sampling when the prior weights of the dead particles are
assigned proportional to
$[(N-1)/N]^i$ at iteration $i$, instead of being based on
Skilling's geometric sequence of $X$-values given by $\exp(-i/N)$.
\citet{walter2017point} demonstrated that the Nested Sampling estimator of $Z$
is unbiased with these weights, but assuming perfect sampling of the replacement
particles at each iteration. Additionally, using the framework of
Sequential Monte Carlo \citep{doucet2001introduction},
\citet{salomone2018unbiased} showed that this remains true even if imperfect
MCMC is used to generate new particles. Technically, this property is
lost if the sequence of distributions is chosen adaptively (a key feature of NS)
rather than being fixed in advanced. However, this paper demonstrates that the
algorithm works well in practice despite this limitation.

@@

{\todo A description of the algorithm for a single rep follows}.
A probability distribution over the objective functions is set up. If there
are only two objective functions, there is just a probability $p_1$ for
objective function 1, and $1-p_1$ for objective function 2.

The first distribution is the prior $\pi(\xx)$. Objective function thresholds
$(\ell_1, \ell_2)_0 = (-\infty, -\infty)$ are assigned.
$N$ particles are generated
from the prior. Then, an objective function is selected according to the
probabilities $\{p_k\}$. Suppose the objective function chosen is number $k$.
Then the worst particle wrt that criteron is written to disk and replaced,
and the threshold is updated in the $k$th position only.

@@

The probability distribution over the objective functions influences the
proportion of the compressions that are done using each objective function.
Different reps ought to have different values for the probability distribution.
Some reps will mostly increase $L_1$, some will mostly increase $L_2$, and
some will be in between.

\section{Example 1: Known solution}
In the software repository, this example is implemented in
{\tt examples/Demo.h} and {\tt examples/Demo.cpp}.
We tested the algorithm on a simple 100-dimensional example, where the
prior is a uniform distribution between 0 and 1 for each coordinate:
\begin{eqnarray}
\pi(\xx) &=& \prod_{i=1}^{100}
\left\{
\begin{array}{lr}
1, & x_i \in [0, 1]\\
0, & \textnormal{otherwise.}
\end{array}
\right.
\end{eqnarray}
The first objective function is quadratic in form, such that the canonical
distributions are iid normal centered at $(0.5, 0.5, ..., 0.5)$:
\begin{eqnarray}
E_1(\xx) &=& \sum_{i=1}^{100} \left(\frac{x_i - 0.5}{0.1}\right)^2.
\end{eqnarray}
The second objective function, given by
\begin{eqnarray} 
E_2(\xx) &=& \sin^2(10 \pi x_i),
\end{eqnarray}
upweights states where coordinates $\{x_i\}$
are close to 0, 0.1, 0.2, ..., 0.9 and 1.
The canonical distributions for this system should draw the coordinates
towards 0.5 (due to $E_1$) and towards the troughs of $E_2$, with the
relative importance of these effects determined by the values of the
inverse temperatures $\beta_1$ and $\beta_2$. Three example canonical
distributions are shown in Figure~\ref{fig:demo_canonical}.
The true partition function (computed numerically) and the Kullback-Leibler
divergence from the prior to the canonical distribution are shown in
Figure~\ref{fig:demo_truth} as a function of two temperatures. These show the
common behaviour where lower temperatures correspond to lower values
of the normalising constant $\log Z$ and higher values of the KL divergence
$H$. There are no phase changes, which would result in discontinuities.

@@

\section{Using the TwinPeaks2018 software}

>>>>

The combination of the reps makes use of the (approximate?) unbiasedness
property of Nested Sampling when the prior weights of the dead particles are
assigned proportional to
$[(N-1)/N]^i$ at iteration $i$, instead of being based on
Skilling's geometric sequence of $X$-values given by $\exp(-i/N)$.
\citet{walter2017point} demonstrated that the Nested Sampling estimator of $Z$
is unbiased with these weights, but assuming perfect sampling of the replacement
particles at each iteration. Additionally, using the framework of
Sequential Monte Carlo \citep{doucet2001introduction},
\citet{salomone2018unbiased} showed that this remains true even if imperfect
MCMC is used to generate new particles. Technically, this property is
lost if the sequence of distributions is chosen adaptively (a key feature of NS)
rather than being fixed in advanced. However, this paper demonstrates that the
algorithm works well in practice despite this limitation.
{\todo Verify/prove statement about bias, and that it means I can average
reps the way I'm doing it. Also check this remains the case with the random
thinning that I'm using.}

@@

{\todo A description of the algorithm for a single rep follows}.
For each rep, we set up a probability distribution over the objective functions,
given by probabilities $\boldsymbol{p} = \{p_1, ..., p_n\}$,
where $n$ is the number of
objective functions. If there
are only two objective functions, there is just a probability $p_1$ for
objective function 1, and $p_2 = 1-p_1$ for objective function 2.

The first distribution is the prior $\pi(\xx)$. Objective function thresholds
$(\ell_1, \ell_2)_0 = (-\infty, -\infty)$ are assigned.
$N$ particles are generated
from the prior. Then, an objective function is selected according to the
probability distribution $\boldsymbol{p}$.
Suppose the objective function chosen is number $k$.
Then the worst particle wrt that criteron is written to disk and replaced,
and the threshold is updated in the $k$th position only.

@@

The probability distribution $\boldsymbol{p}$ influences the
proportion of the compressions that are done using each objective function.
Different reps ought to have different values for the probability distribution.
Some reps will mostly increase $L_1$, some will mostly increase $L_2$, and
some will be in between. In the reference implementation
(Appendix~\ref{sec:software}), the log-probabilities are chosen from a $t$
distribution with 2 degrees of freedom, exponentiated, and then normalized.

\section{Example 1: Known solution}
In the software repository, this example is implemented in
{\tt examples/Demo.h} and {\tt examples/Demo.cpp}.
We tested the algorithm on a simple 100-dimensional example, where the
prior is a uniform distribution between 0 and 1 for each coordinate:
\begin{eqnarray}
\pi(\xx) &=& \prod_{i=1}^{100}
\left\{
\begin{array}{lr}
1, & x_i \in [0, 1]\\
0, & \textnormal{otherwise.}
\end{array}
\right.
\end{eqnarray}
The first objective function is quadratic in form, such that the canonical
distributions are iid normal centered at $(0.5, 0.5, ..., 0.5)$:
\begin{eqnarray}
E_1(\xx) &=& \sum_{i=1}^{100} \left(\frac{x_i - 0.5}{0.1}\right)^2.
\end{eqnarray}
The second objective function, given by
\begin{eqnarr \sum_{i=1}^{100}\sin^2(10 \pi x_i),
\end{eqnarray}
upweights states where coordinates $\{x_i\}$
are close to 0, 0.1, 0.2, ..., 0.9 and 1.
Since both of the objective functions are simple sums over the coordinates,
the canonical distributions are products of independent distributions for
each coordinate.

The canonical distributions for this system should draw the coordinates
towards 0.5 (due to $E_1$) and towards the troughs of $E_2$, with the
relative importance of these effects determined by the values of the
inverse temperatures $\beta_1$ and $\beta_2$. Three example canonical
distributions are shown in Figure~\ref{fig:demo_canonical}.
The true partition function (computed numerically) and the Kullback-Leibler
divergence from the prior to the canonical distribution are shown in
Figure~\ref{fig:demo_truth} as a function of two temperatures. These show the
common behaviour where lower temperatures correspond to lower values
of the normalising constant $\log Z$ and higher values of the KL divergence
$H$. There are no phase changes, which would result in discontinuities.

@@

\section{Using the TwinPeaks2018 software}\label{sec:software}
